---
title: "What do you want to say with your plot?"
subtitle: "Introduction"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
    link-external-newwindow: true
---

## Module overview

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(cowplot)
ggplot2::theme_set(theme_cowplot())
```

1. What do you want to say with your plot?
  - Plot styles
  - Data wrangling
  - Ethics
2. Mapping information onto a plot
3. Constructing complex plots
4. Designing reusable plotting code
5. Finishing touches


## Module prerequisites

- *Quantitative Methods in Life Sciences I*
- Updated R and RStudio
- Basics of data import and manipulation
- Basics of `ggplot`


## Relevant *Quantitative Methods I* course materials

- Week 2: Data organization, loading, visualization
- Week 3: Tidy data and transformations
- [Course files on Github](https://github.com/kmiddleton/quant_methods)
  - Clone to your computer for reference
- [Course videos on YouTube](https://www.youtube.com/playlist?list=PLeIk-G-wZpqx1yW49XAHqoccBXhNJrUSu) 


## Resources

- [R for Data Science](https://r4ds.had.co.nz/)
- [ggplot2](https://ggplot2-book.org/) *In-progress* 3rd edition
- [Modern Data Science with R](https://mdsr-book.github.io/mdsr2e/)
- [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/index.html)
- [Data Visualization: A Practical Introduction](https://socviz.co/index.html#preface)
- [ggplot2 cheatsheet](https://www.maths.usyd.edu.au/u/UG/SM/STAT3022/r/current/Misc/data-visualization-2.1.pdf)
- [Tidyverse style guide](https://style.tidyverse.org/)
- [The Work of Edward Tufte](https://www.edwardtufte.com/tufte/): Data-to-ink ratio


## What we cover and don't cover

Do:

- Basics pretty broadly
- Common cases you will encounter

Don't:

- Every possible use case


## Dual motivations for this module

1. Thinking about what "story" you want to tell with your plots / figures / images
2. Learning about the details of how to show what you want to show.


## Plotting options in R

- Base R
- Lattice
- `ggplot` and the `tidyverse`
  - [ggplot Extensions](https://exts.ggplot2.tidyverse.org/gallery/)


## Packages we will use often

- [tidyverse](https://www.tidyverse.org/) Data wrangling
  - tidyr
  - dplyr
  - etc.
- [cowplot](https://cran.r-project.org/web/packages/cowplot/vignettes/introduction.html) Compositing and refining plots
  - `ggplot2::theme_set(cowplot::theme_cowplot())` is almost always in the setup chunk 
- [patchwork](https://patchwork.data-imaginist.com/) Compositing and refining plots
- [paletteer](https://emilhvitfeldt.github.io/paletteer/) Unifies color palette options


## Why visualize data? {.scrollable}

```{r}
ds <- read_csv("Data/Datasaurus_data.csv",
               col_names = c("x1","x2"),
               show_col_types = FALSE)
knitr::kable(ds)
```


## Why visualize data? {.scrollable}

```{r}
ds |> 
  summarize(across(.cols = everything(),
                   .fns = list(Mean = mean, `Std. Dev.` = sd))) |> 
  mutate(across(.fns = round, digits = 2)) |> 
  knitr::kable()
```

. . .

What is the correlation?

```{r}
#| echo: true
with(ds, cor.test(x1, x2))
```


## Visualization reveals patterns

```{r}
ggplot(ds, aes(x = x1, y = x2)) +
  geom_point(color = "navy") +
  scale_x_continuous(limits = c(0, 100)) +
  scale_y_continuous(limits = c(0, 100)) +
  coord_fixed() +
  labs(title = expression(paste("Alberto Cairo's ",italic("Datasaurus"))))
```


## General rules of plot design

- Some definitely bad options, many possible good designs
- Think about communication
- Personal preference enters


## Presentation-quality plots

Keep in mind throughout this module:

- Plot creation is an iterative process
- Remember what "problems" you have solved before
- Keep a code "gallery" of those solutions
- Mastering ggplot takes time and experience
