---
title: "Styles of plots"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
---

## General caveat to keep in mind

Many ways to accomplish the same goal in R / ggplot.

- Some might be faster than others
  - Unless you are working at large scale, it might not matter
  - Plotting 10^5^ or 10^6^ points? It might.
- Some might be more or less convoluted
  - Pre-compute quantities or use a within-plot function?

Our way might not be the "best" way (for some definition of "best").

## Plots convey information

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(paletteer)
library(cowplot)
ggplot2::theme_set(theme_cowplot())
```

First think about what information you want to convey to yourself or the reader.

## Kinds of data

- Numerical / Quantitative
  - Continuous, discrete
- Categorical / Qualitative
  - Unordered, ordered / nominal

The type of data gives clues about what plots will show, how you set them up, and what elements they include.

## `palmerpenguins`

[https://allisonhorst.github.io/palmerpenguins/](https://allisonhorst.github.io/palmerpenguins/)

```{{r}}
install.packages("palmerpenguins")
```

Look at `penguins`

```{r, echo=TRUE}
library(palmerpenguins)
penguins <- penguins |> 
  mutate(sex = if_else(sex == "female", "Female", "Male"))
names(penguins)
```

## `palmerpenguins` {.smaller}

Mix of categorical (`species`, `island`, `sex`) and continuous (`bill...`, `flipper_length_mm`, `body_mass_g`) variables

- Integer: `year`
- Note: we don't recommend including units in variable names

```{r}
penguins
```

## `palmerpenguins`

Explore:

```{r, echo=TRUE}
penguins |> 
  group_by(species, island, sex) |> 
  count()
```

## A (not exhaustive) menu of plot elements (`geoms`)

- points
- lines (straight, curved)
- bars (histograms)
- densities
- boxes
- annotations

## What information do I want to convey?

- Associations between measurements
- Differences in measurements between species
- Differences in measurements between sexes (within species?)
- Changes in measurements across time

## Associations between measurements

```{r}
penguins |> 
  drop_na(body_mass_g, bill_length_mm, sex) |> 
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point() +
  labs(x = "Body Mass (g)", y = "Bill Length (mm)")
```

## Associations between measurements

```{r}
penguins |> 
  drop_na(body_mass_g, bill_length_mm, sex) |> 
  ggplot(aes(x = body_mass_g, y = bill_length_mm, color = species)) +
  geom_point() +
  labs(x = "Body Mass (g)", y = "Bill Length (mm)") +
  scale_color_paletteer_d(`"beyonce::X56"`, name = "Species")
```

## Associations between measurements

```{r}
penguins |> 
  drop_na(body_mass_g, bill_length_mm, sex) |> 
  ggplot(aes(x = body_mass_g, y = bill_length_mm, color = species)) +
  geom_point() +
  labs(x = "Body Mass (g)", y = "Bill Length (mm)") +
  scale_color_paletteer_d(`"beyonce::X56"`, name = "Species") +
  facet_grid(sex ~ .)
```

## Differences in measurements between species

```{r}
penguins |> 
  drop_na(body_mass_g, sex) |> 
  ggplot(aes(x = species, y = body_mass_g, shape = sex)) +
  geom_point(position = position_jitter()) +
  labs(x = "Species", y = "Body Mass (g)") +
  scale_color_paletteer_d(`"beyonce::X56"`, name = "Species") +
  scale_shape_manual(values = c(1, 16), name = "Sex")
```

## Differences in measurements between sexes within species

```{r}
penguins |> 
  drop_na(body_mass_g, sex) |> 
  ggplot(aes(x = species, y = body_mass_g, color = species,
             group = sex, shape = sex)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.1)) +
  labs(x = "Species", y = "Body Mass (g)") +
  scale_color_paletteer_d(`"beyonce::X56"`, name = "Species") +
  scale_shape_manual(values = c(1, 16), name = "Sex")
```

## Changes in measurements across time

```{r}
penguins |> 
  drop_na(body_mass_g, sex) |> 
  mutate(year = factor(year)) |> 
  ggplot(aes(x = year, y = body_mass_g, color = species)) +
  geom_point(position = position_jitter(width = 0.1)) +
  labs(x = "Year", y = "Body Mass (g)") +
  scale_color_paletteer_d(`"beyonce::X56"`, name = "Species") +
  scale_shape_manual(values = c(1, 16), name = "Sex") +
  facet_grid(species ~ sex)
```

## Plots gone wrong

```{r}
penguins |> 
  drop_na(body_mass_g, sex) |> 
  ggplot(aes(x = species, y = body_mass_g, color = species,
             group = sex, shape = sex)) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge(jitter.width = 0.1)) +
  labs(x = "Species", y = "Body Mass (g)") +
  scale_color_paletteer_d(`"beyonce::X56"`, name = "Species") +
  scale_shape_manual(values = c(1, 16), name = "Sex")
```

