---
title: "Ethics in plotting"
subtitle: "Communication & Deception"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
    link-external-newwindow: true
---

## Responsibility

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(cowplot)
library(colorblindr)
ggplot2::theme_set(theme_cowplot())
```

- The first goal is honest communication
- Accurately and realistically represent the effect (or lack of)
- Match your statistics
- Digestible version often leaves out important details


## Readings

- [Data science ethics](https://mdsr-book.github.io/mdsr2e/ch-ethics.html) in *Modern Data Science with R* by Baumer, Kaplan, and Horton
- [The Ethics of Data Visualization](https://medium.com/@peter.haferl/the-ethical-responsibilities-of-data-visualization-4d12b7c9640d)

## Example

Two variables with a lot of samples. Similar means and standard deviations

```{r}
set.seed(436248)
M <- tibble(x1 = rnorm(1e4, 90, 0.25),
            x2 = rnorm(1e4, 90.5, 0.25))

M

M |> 
  summarize(across(.fns = c(Mean = mean, SD = sd))) |> 
  print(digits = 3)

M <- M |> 
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value")
```

## Two ways to plot the same data

```{r}
p1 <- ggplot(M, aes(Value, color = Variable)) +
  geom_line(stat = "density", size = 1.5) +
  scale_color_OkabeIto() +
  labs(x = "Meaurement", y = "Density",
       title = "Default x Scale") +
  theme(legend.position = c(0.8, 0.8))

p2 <- ggplot(M, aes(Value, color = Variable)) +
  geom_line(stat = "density", size = 1.5) +
  scale_color_OkabeIto() +
  scale_x_continuous(limits = c(80, 100)) +
  labs(x = "Meaurement", y = "Density",
       title = "x Scaled to Biological Range") +
  theme(legend.position = c(0.8, 0.8))
plot_grid(p1, p2, ncol = 2)
```

