---
title: "Repeating plots by a factor"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
    link-external-newwindow: true
---

## Getting more information into a plot

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(paletteer)
library(cowplot)
ggplot2::theme_set(theme_cowplot())

library(Data4Ecologists)
```

- Aesthetics only go so far
  - More than about 2 becomes unwieldy
- Dodging and faceting can provide more options


## Bear movement

Black bear movements in northern Minnesota ^[Ditmer, MA, et a.. 2015. Behavioral and physiological responses of American black bears to landscape features within an agricultural region. *Ecosphere* 6: 1-21.]

```{r}
bear <- bearmove |> 
  rename(log_Move = log.move,
         Heartrate = hr,
         Day = julian) |> 
  select(-BearIDYear) |> 
  mutate(Season = factor(Season),
         DayNight = factor(DayNight),
         Sex = factor(Sex),
         Stage = factor(Stage),
         BearID = factor(BearID)) |> 
  as_tibble()

str(bear)
```


## Aggregate bear movement data {.scrollable}

```{r}
bear |> 
  group_by(BearID, Season, DayNight, Year) |> 
  count() |> 
  knitr::kable()
```


## Dodging points

- Use the `position` argument to shift a `group`.
  - Set `width` by trial and error
- Often used with `color`

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "5-6"
ggplot(bear,
       aes(x = BearID,
           y = log_Move,
           color = DayNight,
           group = DayNight)) +
  geom_point(position = position_dodge(width = 0.2))
```


## Dodging points with jitter

- Use `position = position_jitterdodge`
- Also set `seed` to make the jitter reproducible

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "6-8"
ggplot(bear,
       aes(x = BearID,
           y = log_Move,
           color = DayNight,
           group = DayNight)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15,
                                             dodge.width = 0.3,
                                             seed = 34683))
```


## Dodging points with jitter

Further refinements:

  - Add `alpha` to reduce overplotting

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "9"
ggplot(bear,
       aes(x = BearID,
           y = log_Move,
           color = DayNight,
           group = DayNight)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15,
                                             dodge.width = 0.3,
                                             seed = 34683),
             alpha = 0.25)
```


## Dodging bar graphs

`width = 0.9` leaves no space between bars but leaves space between `BearID`s.

```{r}
#| echo: true
#| output-location: slide
bear |> 
  group_by(BearID, Season) |> 
  summarize(Mean_log_Move = mean(log_Move),
            .groups = "drop") |> 
  ggplot(aes(x = BearID,
             y = Mean_log_Move,
             fill = Season,
             group = Season)) +
  geom_bar(stat = "identity",
           position = position_dodge(width = 0.9))
```


## Faceting

Good for focusing on comparisons among measurements

- Automated generation of multiple plots, each split by the levels of a factor
- By wrapping in factor order with `facet_wrap()`
- By rows or columns with `facet_grid()`, if there is some logical order
  - Think about the direction of the comparison (down rows or across columns) and the way your eyes will move around the plot


## Faceting in rows

`.` is a placeholder for faceting only by row or column

Facet in rows by `Season`:

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "10"
ggplot(bear,
       aes(x = BearID,
           y = log_Move,
           color = DayNight,
           group = DayNight)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15,
                                             dodge.width = 0.3,
                                             seed = 34683),
             alpha = 0.25) +
  facet_grid(Season ~ .)
```


## Faceting in columns

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "10"
ggplot(bear,
       aes(x = BearID,
           y = log_Move,
           color = DayNight,
           group = DayNight)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15,
                                             dodge.width = 0.3,
                                             seed = 34683),
             alpha = 0.25) +
  facet_grid(. ~ Season)
```


## Faceting in rows and columns

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "10"
ggplot(bear,
       aes(x = BearID,
           y = log_Move,
           color = DayNight,
           group = DayNight)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15,
                                             dodge.width = 0.3,
                                             seed = 34683),
             alpha = 0.25) +
  facet_grid(Season ~ DayNight)
```


## Wrapping facets

Number of rows (`nrow`) or columns (`ncol`) can be specified

Wrapping by `BearID` with `facet_wrap()`:

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "10"
ggplot(bear,
       aes(x = Season,
           y = log_Move,
           color = DayNight,
           group = DayNight)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.15,
                                             dodge.width = 0.3,
                                             seed = 34683),
             alpha = 0.25) +
  facet_wrap("BearID")
```


## Axis ranges

- By default, all axis ranges are equal (`fixed`)
  - Usually the behavior you want
- `free_x`, `free_y`, and `free` can be used to allow `x`, `y`, or both axes to vary by facet.
  - Useful for making composite plots in one plot statement
  - Variables with different magnitudes or ranges

## Pivoting the bear data

```{r}
#| echo: true

bear_long <- bear |> 
  select(Season, log_Move, Heartrate, BearID) |> 
  pivot_longer(cols = c(log_Move, Heartrate),
               names_to = "Measurement")
bear_long
```


## Default equal axis scale

```{r}
#| echo: true
#| output-location: slide
ggplot(bear_long,
       aes(x = BearID,
           y = value)) +
  geom_point(position = position_jitter(width = 0.2,
                                        seed = 4527345),
             alpha = 0.25) +
  facet_grid(Measurement ~ Season)
```


## With free axis scale

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "7"
ggplot(bear_long,
       aes(x = BearID,
           y = value)) +
  geom_point(position = position_jitter(width = 0.2,
                                        seed = 4527345),
             alpha = 0.25) +
  facet_grid(Measurement ~ Season, scales = "free_y")
```

