---
title: "Theming: Customizing plots"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
    link-external-newwindow: true
---

## What can you customize?

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(paletteer)
library(cowplot)
ggplot2::theme_set(theme_cowplot())

library(Data4Ecologists)
library(abdData)
```

(Nearly) everything.


## Blank slate

There is a ggplot here:

```{r}
#| echo: true
ggplot() +
  theme_void()
```


## `gg`-this and `gg`-that

- Lots of packages have plotting functionality built on top of `ggplot`
- Not all play nicely with regular `ggplot` theme elements
  - Can't `+` elements, limited ability to modify `theme()` elements, etc.
  - Your mileage may vary


## Scales

`scale_` handles changing how aesthetics are represented in plots

- `scale_x_`
- `scale_y_`
- `scale_color_`
- `scale_fill_`
- `scale_shape_`
- etc.


## Types of scales

`scale_`aesthetic`_`[type]{style="color: #CB3570"} where [type]{style="color: #CB3570"} is

- `continuous`
- `discrete`
- `manual`
- `date`, `time`, `datetime`
- etc.

Different aesthetics have different scale types available

- RStudio tab completion is your friend here: `scale_x`-[Tab]{style="color: #CB3570"}


## `scale_x_` and `scale_y_`

Most commonly:

- `scale_{xy}_continuous`: Continuous
- `scale_{xy}_discrete`: Categorical
- `scale_{xy}_log10`: log~10~ transformation
- `scale_{xy}_sqrt`: Square-root transformation
- `scale_{xy}_reverse`: Reverse the axis


## Electric fish diversity and tributaries {.scrollable}

Species abundance of electric fish upstream and downstream of the entrance of a tributary in the Amazon basin.

```{r}
EF <- ElectricFish |> 
  rename(Tributary = tributary,
         n_Upstream = species.upstream,
         n_Downstream = species.downstream) |> 
  arrange(Tributary)
EF |> knitr::kable()
```


## Electric fish visualization

1. Bar plot of `n_Upstream` and `n_Downstream` diversity by `Tributary`
2. Scatter plot of `n_Upstream` vs. `n_Downstream`


## Bar plot

```{r}
#| echo: true
#| output-location: slide
EF_bar <- EF |> 
  pivot_longer(cols = -Tributary,
               names_to = "Up_Down",
               values_to = "n_Species") |> 
  ggplot(aes(x = Tributary,
             y = n_Species,
             fill = Up_Down,
             group = Up_Down)) +
  geom_bar(stat = "identity",
           position = position_dodge(width = 0.9))
EF_bar
```


## Bar plot refinements

- `ggplot`s are additive, so we can "add" to an existing plot.
- Change axis limits (but see *Ethics in visualization*)
- Two ways to change axis labels

```{r}
#| echo: true
#| output-location: slide
EF_bar +
  scale_y_continuous(limits = c(0, 50),
                     breaks = seq(0, 50, by = 5)) +
  scale_x_discrete(name = "Upstream/Downstream") +
  labs(y = "Number of Species",
       title = "Species Diversity in the Amazon")
```


## Scatter plot

What is the relationship between upstream and downstream diversity?

```{r}
#| echo: true
#| output-location: slide
ggplot(EF, aes(x = n_Upstream,
               y = n_Downstream)) +
  geom_point()
```


## Scatter plot refinements

- Add a straight line with a slope of 1
- Change the color and size of all points

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "3-4"
ggplot(EF, aes(x = n_Upstream,
               y = n_Downstream)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_point(color = "tomato", size = 3)
```


## Changing axis aspect ratio

- Make the plot square (`coord_equal`)
- Set the axis ranges to be equal (`xlim`, `ylim`)

```{r}
#| echo: true
#| output-location: slide
ggplot(EF, aes(x = n_Upstream,
               y = n_Downstream)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_point(color = "tomato", size = 3) +
  coord_equal(xlim = c(0, 30), ylim = c(0, 30))
```


## `scale_color_` and `scale_fill_`

- Set the fill manually
- Include a line break in the legend title string with `\n`

```{r}
#| echo: true
#| output-location: slide
EF_bar +
  scale_fill_manual(values = c("orange", "darkgreen"),
                    name = "Upstream /\nDownstream")
```


## Automated color and fill scales

- [RColorBrewer](https://colorbrewer2.org/)
- [viridis](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html)
- [paletteer](https://emilhvitfeldt.github.io/paletteer/)

```{r}
#| echo: true
#| output-location: slide
EF_bar +
  scale_fill_brewer(type = "qual")
```


## `scale_shape_`

- ggplot inherits the standard set of R point shapes
- [Point shapes in R](https://r-graphics.org/recipe-scatter-shapes)


## `scale_size_`

- [Size by area (default) or radius](https://ggplot2-book.org/scale-other.html)
- Set the `range` to control the output scale

```{r}
#| echo: true
#| output-location: slide
#| code-line-numbers: "5-11"
EF_size <- EF |> 
  pivot_longer(cols = -Tributary,
               names_to = "Up_Down",
               values_to = "n_Species") |> 
  ggplot(aes(x = Tributary,
             y = Up_Down,
             label = n_Species,
             size = n_Species)) +
  geom_point(alpha = 0.25) +
  geom_text() +
  scale_size(range = c(1, 10))
EF_size
```


## `theme()` changes aspects of plots

- Fully custom and reusing themes in a few weeks
- See `?theme` for options
- [ggplot book has many examples](https://ggplot2-book.org/polishing.html)

```{r}
EF_bar
```


## Changing axis components
```{r}
#| echo: true
#| output-location: slide
EF_bar +
  labs(x = "Stream") +
  scale_fill_manual(values = c("darkred", "navy"),
                    name = "Location",
                    labels = c("Downstream", "Upstream")) +
  theme(axis.text = element_text(size = 9),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        axis.title.y = element_text(face = "bold"),
        axis.title.x = element_blank(),
        legend.title = element_text(size = 10, face = "italic"),
        legend.text = element_text(size = 9))
```


## (Re)moving and reshaping the legend

- Remove the legend

```{r}
#| echo: true
EF_size +
  theme(legend.position = "none")
```


## Customizing the legend

```{r}
#| echo: true
#| output-location: slide
ggplot(EF, aes(x = n_Upstream,
               y = n_Downstream,
               color = Tributary)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_point(size = 5) +
  scale_color_viridis_d(guide = guide_legend(ncol = 3,
                                             override.aes = list(size = 2))) +
  coord_equal(xlim = c(0, 30), ylim = c(0, 30)) +
  labs(x = "Upstream Species",
       y = "Downstream Species",
       title = "Species Diversity in the Amazon") +
  theme(legend.title = element_text(size = 10),
        legend.text = element_text(size = 9),
        legend.position = c(0.35, 0.175),
        legend.box.background = element_rect(color = "black", size = 1),
        legend.box.margin = margin(4, 4, 4, 4),
        axis.title = element_text(face = "bold"))
```


## Using `element_blank()` to remove plot elements

- Remove the `x`-axis completely
- Remove the space between the labels and bars

```{r}
#| echo: true
#| output-location: slide
#| message: false
EF_bar +
  scale_y_continuous(expand = c(0, 0)) +
  theme(axis.ticks.x = element_blank(),
        axis.line.x = element_blank())
```

