---
title: "Theming: Customizing plots"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
---

## What can you customize?

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(paletteer)
library(cowplot)
ggplot2::theme_set(theme_cowplot())

library(Data4Ecologists)
library(abdData)
```

(Nearly) everything.

## Blank slate

```{r}
#| echo: true
ggplot() +
  theme_void()
```

## Scales

`scale_` handles changing how aesthetics are represented in plots

- `scale_x_`
- `scale_y_`
- `scale_color_`
- `scale_fill_`
- `scale_shape_`
- etc.

## Types of scales

`scale_`aesthetic`_`[type]{style="color: #CB3570"} where [type]{style="color: #CB3570"} is

- `continuous`
- `discrete`
- `manual`
- `datetime`
- etc.

Different aesthetics have different scale types available

- RStudio tab completion is your friend here: `scale_x`-[Tab]{style="color: #CB3570"}

## `scale_x_` and `scale_y_`

## Electric fish diversity and tributaries {.scrollable}

Species abundance of electric fish upstream and downstream of the entrance of a tributary in the Amazon basin.

```{r}
EF <- ElectricFish |> 
  rename(Tributary = tributary,
         n_Upstream = species.upstream,
         n_Downstream = species.downstream) |> 
  arrange(Tributary)
EF |> knitr::kable()
```

## Electric fish visualization

1. Bar plot of Upstream and Downstream by Tributary
2. Scatterplot of Upstream vs. Downstream

## `scale_color_`

## `scale_fill_`

## `scale_shape_`

## `scale_size_`

## `theme()` changes aspects of plots

- Custom and reusable themes in a few weeks

## Changing axis components

## Changing axis aspect ratio

## (Re)moving and reshaping the legend
