---
title: "Constructing Complex Plots"
subtitle: "Combining Data Sources"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
    link-external-newwindow: true
---

## This week

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(cowplot)
ggplot2::theme_set(theme_cowplot())

library(Data4Ecologists)
library(abdData)
```

1. Combining data sources
2. Adding elements
3. Highlighting elements
4. Reusable themes


## "Complex plots"

- Combining `geom`s
- Data from different sources (`data.frame`s, regressions, etc.)
- Non-standard elements or axis labels (text with math symbols)
- Custom colors or gradients
- Highlighting different elements with color, labels or arrows


## Zooplankton diversity {.scrollable}

Diversity of zooplankton (`Diversity`) prey in each of 5 replicate `Block`s of three `Treatment` levels.^[Svanbäck, R and DI Bolnick. 2007. Intraspecific competition drives increased resource use diversity within a natural population. *Proc R Soc London Ser B, Biol Sci* 274: 839-844.]

```{r}
ZP <- Zooplankton |> 
  rename(Treatment = treatment,
         Diversity = zooplankton,
         Block = block) |> 
  mutate(Treatment = case_when(
    Treatment == "control" ~ "Control",
    Treatment == "low" ~ "Low",
    Treatment == "high" ~ "High"
  ),
  Treatment = factor(Treatment, levels = c("Low", "Control", "High")),
  Block = factor(Block)) |> 
  as_tibble()

ZP
```


## Points by groups

::: {.panel-tabset}

### Code

```{r}
#| echo: true
p <- ggplot(ZP, aes(x = Treatment, y = Diversity)) +
  geom_point(size = 2,
             position = position_jitter(width = 0.05,
                                        seed = 6734747))
```

### Plot

```{r}
p
```

:::


## Boxplot

::: {.panel-tabset}

### Code

```{r}
#| echo: true
p <- ggplot(ZP, aes(x = Treatment, y = Diversity)) +
  geom_boxplot()
```

### Plot

```{r}
p
```

:::


## Boxplot with points

::: {.panel-tabset}

### Code

- Add points on top of the boxplot (`ggplot` adds `geom`s sequentially)
- Turn off automatic outlier identification (`outlier.shape = NA`)

```{r}
#| echo: true
p <- ggplot(ZP, aes(x = Treatment, y = Diversity)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(size = 2,
             position = position_jitter(width = 0.05,
                                        seed = 6734747))
```

### Plot

```{r}
p
```

:::


## Precomputing quantities

```{r}
#| echo: true
SEM <- function(x) {
  return(sd(x) / sqrt(length(x)))
}

ZP_means <- ZP |> 
  group_by(Treatment) |> 
  summarise(mean_Diversity = mean(Diversity),
            upper_SEM = mean_Diversity + 2 * SEM(Diversity),
            lower_SEM = mean_Diversity - 2 * SEM(Diversity),
            .groups = "drop")

ZP_means
```


## Plots from multiple data sources


::: {.panel-tabset}

### Code

- Use `data` argument to `geom_`...

```{r}
#| echo: true
p <- ggplot() +
  geom_point(data = ZP, aes(x = Treatment, y = Diversity),
             size = 2,
             position = position_jitter(width = 0.05,
                                        seed = 6734747)) +
  geom_point(data = ZP_means,
             aes(x = Treatment, y = mean_Diversity),
             size = 4,
             color = "gray50") +
  geom_errorbar(data = ZP_means,
                aes(x = Treatment, ymin = lower_SEM, ymax = upper_SEM),
                width = 0.1,
                size = 1,
                color = "gray50")
```

### Plot

```{r}
p
```

:::


## Reaction norms for paired data

Antibody levels in Red-winged blackbirds before and after experimental manipulation of testosterone^[Hasselquist D, JA Marsh, PW Sherman, JC Wingfield. 1999. Is avian immunocompetence suppressed by testosterone? *Behav Ecol Sociobiol* 45: 167-175.]

```{r}
BB <- Blackbirds |> 
  mutate(ID = factor(1:n())) |> 
  rename(Before = before,
         After = after) |> 
  select(ID, Before, After)

BB
```


## Pivoting

```{r}
#| echo: true
BB_long <- BB |> 
  pivot_longer(cols = -ID,
               names_to = "Timepoint",
               values_to = "Antibody_level")

BB_long
```


## Plotting reaction norms

::: {.panel-tabset}

### Code

```{r}
#| echo: true
p <- ggplot(BB_long, aes(x = Timepoint, y = Antibody_level, group = ID)) +
  geom_point() +
  geom_line()
```

### Plot

```{r}
p
```

:::


## Ordering factors

::: {.panel-tabset}

### Code

Also many options in the `forcats` package (`fct_` functions) when you already have a factor.

```{r}
#| echo: true

BB_long <- BB_long |> 
  mutate(Timepoint = factor(Timepoint, levels = c("Before", "After")))

p <- ggplot(BB_long, aes(x = Timepoint, y = Antibody_level, group = ID)) +
  geom_point() +
  geom_line()
```

### Plot

```{r}
p
```

:::


## Adding a mean and error bars

::: {.panel-tabset}

### Code

- `stat_summary()` can add the output of summary functions.
- `mean_cl_boot` give the mean and bootstrapped 95% CI. (requires `Hmisc`)
- `aes(group = -1)` ungroups the data so that the statistics will calculate for each `Timepoint`

```{r}
#| echo: true
p <- ggplot(BB_long, aes(x = Timepoint, y = Antibody_level, group = ID)) +
  geom_point() +
  geom_line() +
  stat_summary(aes(group = -1),
               fun.data = "mean_cl_boot",
               color = "goldenrod",
               size = 1)
```

### Plot

```{r}
p
```

:::

## Horizontal, Vertical, and `ab` lines


## Adding regression lines


## `geom_smooth()`


## `predict()` with `interval="predict"`

- Grid construction with `crossing()`


## `fortify()`


## Making confidence interval bands


