---
title: "Adding elements"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
    link-external-newwindow: true
---

## Data-to-ink ratio

Don’t add complexity to your plot without a good reason.

Some journals don't like text in plots.

## Annotating plots

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(cowplot)
ggplot2::theme_set(theme_cowplot())
library(paletteer)
library(lubridate)
library(png)
library(ggpubr)
library(ggmap)

library(Data4Ecologists)
library(abdData)
```

Adding elements to plots

- Lines
- Boxes, shading plot areas
- Images (jpg, png)
- Maps
- Text and math text


## `annotate()`

- Adds elements without using `geom_`
- Doesn't require `data.frames` or `tibbles`
- Useful for single elements


## Boxes and shading

- Time series (day/night, seasons, years)
- Differentiate chromosomes, etc.

`geom_rect()` uses values for four corners: `xmin`, `xmax`, `ymin`, and `ymax`


## Mouse wheel activity

```{r}
#| echo: true
WR <- read_csv("Data/Mouse_wheel.csv", show_col_types = FALSE) |> 
  mutate(Date = mdy_hm(Date) - hours(1)) |> 
  pivot_longer(cols = -Date, names_to = "Wheel", values_to = "Revolutions") |> 
  mutate(Wheel = str_remove(Wheel, "Wheel") |> as.integer()) |> 
  filter(Wheel <= 4)

WR
```

## Create a tibble with `x` bounds of the box

```{r}
#| echo: true
DayNight <- tibble(
  Lights_off = ymd_hms(c("2013-05-07 19:00:00", "2013-05-08 19:00:00")),
  Lights_on = ymd_hms(c("2013-05-08 07:00:00", "2013-05-09 07:00:00"))
)
```

```{r}
DayNight
```

## Plot with background `geom_rect()`

- `Inf` and `-Inf` used when you don't know the limits

```{r}
#| echo: true
#| output-location: slide
ggplot() +
  geom_rect(data = DayNight,
            aes(xmin = Lights_off, xmax = Lights_on,
                ymin = 0, ymax = Inf),
            fill = "gray80") +
  geom_path(data = WR, aes(x = Date, y = Revolutions, group = Wheel)) +
  facet_wrap("Wheel")
```


## Refine plot

- Change date labels, rotate text
- Change background of facet headers

```{r}
#| echo: true
#| output-location: slide
ggplot() +
  geom_rect(data = DayNight,
            aes(xmin = Lights_on, xmax = Lights_off,
                ymin = 0, ymax = Inf),
            fill = "gray80") +
  geom_path(data = WR, aes(x = Date, y = Revolutions, group = Wheel)) +
  facet_wrap("Wheel") +
  scale_x_datetime(date_labels = "%H:%M",
                   name = "Time") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        strip.background = element_rect(fill = "goldenrod"))
```


## Overlaying data on Images

- Loading image
- Setting up coordinates with `tribble()`

```{r}
#| echo: true
img <- png::readPNG("Images/Craniometric_tracing.png")

coords <- tibble::tribble(
  ~ LM, ~ x, ~ y,
  "ANS", 318, 249,
  "Articulare", 44, 193,
  "Basion", 7, 221,
  "Condylion", 59, 169,
  "Gonion", 73.5, 332,
  "Menton", 215, 415,
  "Nasion", 313, 96,
  "PNS", 153, 241,
  "Pogonion", 239, 424,
  "Sella", 79, 116
) |> 
  mutate(y = dim(img)[1] - y) # Reverse due to R/ImageJ differences
```


## Images are matrices (grayscale) or arrays (color)

```{r}
#| echo: true
class(img)
str(img)
img[1:10, 1:10]
```


## plot with `annotation_raster()`

- `xmin`, `xmax`, `ymin`, and `ymax` define the box for the image
- `dim(img)` gives the dimensions of the matrix
- Use `ggpubr::theme_transparent()` to remove all plot axes, labels, etc.

```{r}
#| echo: true
#| output-location: slide
ggplot(data = coords, aes(x, y)) +
    annotation_raster(img,
                      xmin = 0, xmax = dim(img)[2],
                      ymin = 0, ymax = dim(img)[1]) +
    geom_point(color = "red", size = 5) +
    xlim(c(0, dim(img)[2])) +
    ylim(c(0, dim(img)[1])) +
    coord_equal() +
    ggpubr::theme_transparent()
```

## Maps

Field sites along the Missouri River

```{r}
#| echo: true
library(ggmap)

Locations <- 
  tribble(~ lat,     ~lon,       ~ ID,
          38.666710, -91.867078, "Mokane",
          38.587094, -92.176563, "Jefferson City",
          38.688121, -92.363351, "Marion",
          38.799781, -92.378246, "Easley",
          38.974020, -92.565014, "Rocheport",
          38.980450, -92.740881, "Boonville",
          39.064000, -92.935129, "Arrow Rock",
          39.232537, -92.848955, "Glasgow")
```


## Setup and plot map with data

```{r}
#| echo: true
#| output-location: slide
#| cache: true

# Create bounding box for locations
sbbox <- make_bbox(lon = Locations$lon,
                   lat = Locations$lat,
                   f = 0.25) # Expansion factor

# Download map
sq_map <- get_stamenmap(bbox = sbbox,
                        maptype = "terrain")

p <- ggmap(sq_map) + 
  geom_point(data = Locations,
             mapping = aes(x = lon, y = lat),
             color = "red",
             size = 3) +
  labs(x = "Longitude", y = "Latitude")
p
```


## Text annotations

```{r}
#| echo: true
#| output-location: slide
p +
  annotate(geom = "text",
           x = -92.45,
           y = 39.35,
           label = "Field Sites on the Missouri River",
           size = 8)
```


## Labeling points with `ggrepel` package

```{r}
#| echo: true
#| output-location: slide
#| warning: false
library(ggrepel)

p +
  annotate(geom = "text",
           x = -92.45,
           y = 39.35,
           label = "Field Sites on the Missouri River",
           size = 8) +
  geom_text_repel(data = Locations,
                  mapping = aes(x = lon, y = lat, label = ID),
                  box.padding = 1,
                  seed = 431579)
```


## Working with math

Several options

- `?plotmath` which uses `expression()` and `paste()`
- `annotate()` with `parse = TRUE`
- `latex2exp` [package](https://cran.r-project.org/web/packages/latex2exp/vignettes/using-latex2exp.html) converts LaTeX expressions to [plotmath](https://stat.ethz.ch/R-manual/R-patched/library/grDevices/html/plotmath.html)


## Regression data

```{r}
set.seed(43577)
M <- tibble(
  x = runif(20, 0, 10),
  y = x * 4 + rnorm(20, 0, 2)
)
M
```


## Modifying axis titles

Note:

- `~` for intentional space, other spaces are collapsed
- No `"`

```{r}
#| echo: true
#| output-location: slide
ggplot(M, aes(x, y)) +
  geom_point() +
  labs(x = expression(Area~(sqrt(m^2))),
       y = expression(Species~density~(Species / m^2)))
```



## Adding an annotation

```{r}
#| echo: true
#| output-location: slide
fm <- lm(y ~ x, data = M)
s <- summary(fm)

ggplot(M, aes(x, y)) +
  geom_point() +
  labs(x = expression(Area~(sqrt(m^2))),
       y = expression(Species~density~(Species / m^2))) +
  annotate(geom = "text",
           label = paste("italic(R) ^ 2 == ",
                         round(s$r.squared, 2)),
           parse = TRUE,
           x = 7.5, y = 10,
           size = 5)

```


## Formatting text with `Tex()`

- Uses raw strings: `TeX(r'($ code goes here $)')`
- [LaTeX math](https://en.wikibooks.org/wiki/LaTeX/Mathematics)
- Works a little differently in `annotate()`

```{r}
#| echo: true
#| output-location: slide
library(latex2exp)

ggplot(M, aes(x, y)) +
  geom_point() +
  labs(x = TeX(r'($ \sqrt{\frac{1}{Area}} \cdot \cos (2\pi) + \theta_{1}$)'),
       y = TeX(r'(Species density$\,\,\left( \frac{Spp.}{m^2} \right) $)')) +
  annotate(geom = "text",
           label = TeX(paste('$\\textit{R}^2 =$', round(s$r.squared, 2)),
                       output = "character"),
           parse = TRUE,
           x = 7.5, y = 10,
           size = 5)

```
