---
title: "Creating reusable themes"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
    link-external-newwindow: true
---

## Why reusable themes?

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(cowplot)
library(ggforce)
ggplot2::theme_set(theme_cowplot())

library(Data4Ecologists)
```

- Many plots share a `scale_color`, `scale_fill`, `scale_linetype`, or `scale_shape`
- Many plots get the same axis text or title treatment
  - Font family, face, size, color
  - Useful for making plots for presentations (powerpoint/equivalent or poster)
- Save time for your future self


## Trunks data

- DBH and trunk flare for 4 species of maple

```{r}
# Setup trunkfl data
Trunks <- trunkfl |> 
  mutate(Species =  case_when(
    Species == 2L ~ "Acer platanoides",
    Species == 3L ~ "Acer rubrum",
    Species == 4L ~ "Acer saccharinum", 
    Species == 5L ~ "Acer saccarum",
    Species == 6L ~ "Fraxinus americana",
    Species == 7L ~ "Fraxinus pennslyvanica"),
    City = case_when(
      City == 1L ~ "Crookston",
      City == 2L ~ "Hendricks",
      City == 3L ~ "Hibbing",
      City == 4L ~ "Hutchinson",
      City == 5L ~ "Morris",
      City == 6L ~ "Rochester"
    )) |> 
  select(Species, City, DBH, TF) |> 
  filter(DBH < 60, str_detect(Species, "Acer")) |> 
  as_tibble()
Trunks
```


## Violin & Sina plot

```{r}
#| echo: true
#| output-location: slide
ggplot(Trunks, aes(x = Species, y = DBH, color = Species)) +
  geom_violin() +
  geom_sina(seed = 74645) +
  scale_color_manual(values = c("red", "orange", "blue", "purple"),
                     breaks = c("Acer saccharinum", "Acer rubrum",
                                "Acer platanoides", "Acer saccarum")) +
  theme(axis.title = element_text(face = "bold"),
        axis.text.x = element_text(face = "italic")) +
  guides(color = "none")
```


## Creating your own reusable theme

- In `ggplot`, elements are added with `+`
- Convert to a list, separate items with `,`

```{r}
#| echo: true
#| output-location: slide
mytheme <- list(
  scale_color_manual(values = c("red", "orange", "blue", "purple"),
                     breaks = c("Acer saccharinum", "Acer rubrum",
                                "Acer platanoides", "Acer saccarum")),
  theme(axis.title = element_text(face = "bold"))
)

ggplot(Trunks, aes(x = Species, y = DBH, color = Species)) +
  geom_violin() +
  geom_sina(seed = 74645) +
  mytheme +
  guides(color = "none") +
  theme(axis.text.x = element_text(face = "italic"))
```


## Reuse your theme

```{r}
#| echo: true
#| output-location: slide
ggplot(Trunks, aes(x = Species, y = TF, color = Species)) +
  geom_violin() +
  geom_sina(seed = 74645) +
  mytheme +
  guides(color = "none") +
  theme(axis.text.x = element_text(face = "italic"))
```


## Reuse your theme

```{r}
#| echo: true
#| output-location: slide
ggplot(Trunks, aes(x = DBH, y = TF, color = Species)) +
  geom_point() +
  mytheme +
  guides(color = guide_legend(label.theme = element_text(face = "italic"))) +
  theme(legend.position = c(0.7, 0.2))
```

