---
title: "Problem Set 2"
author:
  - Your Name Here
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    toc-title: Contents
---

```{r}
#| echo: false
#| message: false

library(tidyverse)
library(cowplot)

theme_set(theme_cowplot())

# Required files for this problem set:

```

# Randomization test

We will set up a simple randomization test of the means below using a `for` loop. In Quant Methods 1, we did many similar for loops. We also showed how to search through the object returned by `summary()` to find the relevant test statistic. Instead, here we will use the simpler tibble returned by `broom::tidy()`. Set `eval` to `true` before you knit.

```{r}
#| eval: true
#| echo: true

MM <- read_csv("../Data/small_data.csv", show_col_types = FALSE)

ggplot(MM, aes(x = Group, y = y)) +
  geom_point()

fm <- lm(y ~ Group, data = MM)

# FIXME
library(broom)

iters <- 1000
t_stats <- numeric(iters)
t_stats[1] <- as.numeric(tidy(fm)[2, 4])

for (ii in 2:iters) {
  MM_rand <- MM |> mutate(Group = sample(Group))
  fm_summary <- lm(y ~ Group, data = MM_rand) |> 
    summary() |> 
    tidy()
  t_stats[ii] <- as.numeric(fm_summary[2, 4])
}

ggplot(data = tibble(t_stats),
       aes(t_stats)) +
  geom_histogram(bins = 50, fill = "gray75") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  geom_vline(xintercept = t_stats[1], color = "dodgerblue4", linewidth = 1)

mean(t_stats[1] >= t_stats)
```



```{r}
#| echo: true

d_obs <- MM |> 
    group_by(Group) |> 
    summarize(y_mean = mean(y)) |> 
    pivot_wider(names_from = Group, values_from = y_mean) |> 
    mutate(d = Group_2 - Group_1) |> 
    pull(d)

# FIXME
Group_diff <- function(.x, MM) {
  MM_resample <- MM |> 
    mutate(Group = sample(Group))
  d <- MM_resample |> 
    group_by(Group) |> 
    summarize(y_bar = mean(y)) |> 
    pivot_wider(names_from = Group, values_from = y_bar) |> 
    mutate(d = Group_2 - Group_1) |> 
    pull(d)
  return(d)
}


# With map
niter <- 1000
d_resample <- map_dbl(.x = seq_len(niter),
                      .f = Group_diff,
                      MM = MM)
d_resample[1] <- d_obs

ggplot(data = tibble(d_resample),
       aes(d_resample)) +
  geom_histogram(bins = 50, fill = "gray75") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  geom_vline(xintercept = d_obs, color = "dodgerblue4", linewidth = 1)

mean(d_resample <= d_obs)
```

How many iterations?

```{r}
choose(3, 2)
```

1. 1 and 2
2. 1 and 3
3. 2 and 3

```{r}
choose(10, 5)

1 / choose(10, 5)

length(unique(d_resample))
```

Repeat for `large_data.csv`

```{r}

```

```{r}
choose(50, 25)
```


## Prove the central limit theorem

```{r}
SP <- abdData::SticklebackPlates

ggplot(SP, aes(plates)) +
  geom_histogram(bins = 30) +
  facet_grid(genotype ~ .)

SP |> 
  group_by(genotype) |> 
  summarize(Mean = mean(plates),
            Median = median(plates))

Mm <- SP |> 
  filter(genotype == "Mm") |> 
  pull(plates)

resample_Mm <- function(ii, nsamp) {
  ids <- sample(seq_len(length(Mm)), nsamp)
  return(tibble(ii, bar = mean(Mm[ids])))
}

niter <- 1000
resamples <- map(.x = seq_len(niter),
                 .f = resample_Mm,
                 nsamp = 20) |> 
  list_rbind()

ggplot(resamples, aes(bar)) +
  geom_histogram(bins = 30)

mean(resamples$bar)
median(Mm)

```


```{r}
library(boot)

mean_fun <- function(data, indices) {
  data_sub <- data[indices]
  return(mean(data_sub))
}

b <- boot(data = Mm, statistic = mean_fun, R = 2000, stype = "i")
b
plot(b)
boot.ci(b, type="bca")
```


Allele frequency

```{r}

freqs <- rbinom(20, 50, 0.2)/50

SNP_A1 <- tibble("Allele" = rbinom(50*20, rep(1, 50*20), freqs),
                 "Location" = paste0("L",rep(1:20, each=50)))

pA1.obs <- mean(SNP_A1$Allele)

jk.PA1 <- tibble("ps" = rep(NA, length=nrow(SNP_A1)))

for(ii in 1:nrow(SNP_A1))
{
  
  pA1.est <- mean(SNP_A1$Allele[-ii])
  jk.PA1$ps[ii] <- nrow(SNP_A1)*pA1.obs - (nrow(SNP_A1)-1)*pA1.est
}


```


An example interval test

jackknife
bootstrap

compare performance for simulated vals? - later?


A few randomizations
