---
title: "Priors and Posteriors"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    embed-resources: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
    link-external-newwindow: true
bibliography: Bayes.bib
server: shiny
---

## Priors

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(lubridate)
library(cowplot)
ggplot2::theme_set(theme_cowplot())
library(patchwork)

library(Data4Ecologists)
library(rethinking)
```

```{r}
Carp <- carppassage |> 
  filter(Sound == 0) |> 
  mutate(Date = mdy(Date)) |> 
  arrange(Date) |> 
  select(Pass, FishID, Date)
```

Priors let us model with knowledge included:

- Almost no knowledge
  - Carp either pass or do not pass
- A little knowledge
  - We guess that the proportion is probably less than 50%
- A lot of knowledge
  - Previous studies show that <25% of carp pass


## Distributions of proportions (and probabilities)

- Bounded by 0 and 1
- Continuous (any proportion is possible)

. . .

The **Beta distribution** fits these requirements

- Two parameters
  - a, $\alpha$, `shape1` in `dbeta()`
  - b, $\beta$, `shape2` in `dbeta()`


## Beta distribution: a = 1, b = 1

- `dbeta(x, shape1, shape2)` is the density of the Beta distribution for a vector `x` and the two shape parameters
- Create an evenly spaced sequence of `P` probabilities from 0 to 1
- Calculate the density for each `P`, given `shape1` and `shape2`

```{r}
#| echo: true
#| output-location: slide
ggplot(data = tibble(P = seq(0, 1, length.out = 200),
                     Density = dbeta(P, shape1 = 1, shape2 = 1)),
       aes(P, Density)) +
  geom_line()
```


## Beta distribution: a = 2, b = 2

```{r}
#| echo: true
#| output-location: slide
ggplot(data = tibble(P = seq(0, 1, length.out = 200),
                     Density = dbeta(P, shape1 = 2, shape2 = 2)),
       aes(P, Density)) +
  geom_line()
```


## Interactive Beta distribution 

::: {layout-ncol="3"}

```{r}
sliderInput("a", "a:", 
            min = 1, max = 20, value = 1)
```

```{r}
sliderInput("b", "b:", 
            min = 1, max = 20, value = 1)
```

:::

```{r}
plotOutput("distPlot")
```

```{r}
#| context: server
output$distPlot <- renderPlot({
  ggplot(data = tibble(P = seq(0, 1, length.out = 200),
                       Density = dbeta(P, input$a, input$b)),
         aes(P, Density)) +
    geom_line()
})
```


## Beta distribution: a = 0.5, b = 0.5

```{r}
ggplot(data = tibble(P = seq(0, 1, length.out = 200),
                     Density = dbeta(P, 0.5, 0.5)),
       aes(P, Density)) +
  geom_line()
```


## Beta distribution for carp observations

- Start with a Beta(1, 1) *prior*
  - All we know is that carp either Pass or Do Not Pass

```{r}
ggplot(data = tibble(P = seq(0, 1, length.out = 200),
                     Density = dbeta(P, 1, 1)),
       aes(P, Density)) +
  geom_line()
```


## Begin observing carp

```{r}
head(Carp)
```

- If `Pass` is 1: add 1 to `a`
- If `Pass` is 0: add 1 to `b`


## Carp 1: No pass

```{r}
Prior <- tibble(Pr_Post = "Prior",
                P = seq(0, 1, length.out = 200),
                a = 1,
                b = 1,
                Density = dbeta(P, a, b))
Posterior <- tibble(Pr_Post = "Posterior",
                P = seq(0, 1, length.out = 200),
                a = 1,
                b = 2,
                Density = dbeta(P, a, b))
D <- bind_rows(Prior, Posterior) |> 
  mutate(Pr_Post = factor(Pr_Post, levels = c("Prior", "Posterior")))

ggplot(D, aes(P, Density, color = Pr_Post)) +
  geom_line() +
  scale_color_manual(values = c("gray50", "red"), name = NULL) +
  labs(title = "a = 1, b = 2")
```


## Carp 2: No pass

```{r}
Prior <- tibble(Pr_Post = "Prior",
                P = seq(0, 1, length.out = 200),
                a = 1,
                b = 2,
                Density = dbeta(P, a, b))
Posterior <- tibble(Pr_Post = "Posterior",
                P = seq(0, 1, length.out = 200),
                a = 1,
                b = 3,
                Density = dbeta(P, a, b))
D <- bind_rows(Prior, Posterior) |> 
  mutate(Pr_Post = factor(Pr_Post, levels = c("Prior", "Posterior")))

ggplot(D, aes(P, Density, color = Pr_Post)) +
  geom_line() +
  scale_color_manual(values = c("gray50", "red"), name = NULL) +
  labs(title = "a = 1, b = 3")
```


## Carp 3: No pass

```{r}
Prior <- tibble(Pr_Post = "Prior",
                P = seq(0, 1, length.out = 200),
                a = 1,
                b = 3,
                Density = dbeta(P, a, b))
Posterior <- tibble(Pr_Post = "Posterior",
                P = seq(0, 1, length.out = 200),
                a = 1,
                b = 4,
                Density = dbeta(P, a, b))
D <- bind_rows(Prior, Posterior) |> 
  mutate(Pr_Post = factor(Pr_Post, levels = c("Prior", "Posterior")))

ggplot(D, aes(P, Density, color = Pr_Post)) +
  geom_line() +
  scale_color_manual(values = c("gray50", "red"), name = NULL) +
  labs(title = "a = 1, b = 4")
```


## Carp 7: Pass

```{r}
Prior <- tibble(Pr_Post = "Prior",
                P = seq(0, 1, length.out = 200),
                a = 1,
                b = 7,
                Density = dbeta(P, a, b))
Posterior <- tibble(Pr_Post = "Posterior",
                P = seq(0, 1, length.out = 200),
                a = 2,
                b = 7,
                Density = dbeta(P, a, b))
D <- bind_rows(Prior, Posterior) |> 
  mutate(Pr_Post = factor(Pr_Post, levels = c("Prior", "Posterior")))

ggplot(D, aes(P, Density, color = Pr_Post)) +
  geom_line() +
  scale_color_manual(values = c("gray50", "red"), name = NULL) +
  labs(title = "a = 2, b = 7")
```


## Carp 8: No pass

```{r}
Prior <- tibble(Pr_Post = "Prior",
                P = seq(0, 1, length.out = 200),
                a = 2,
                b = 7,
                Density = dbeta(P, a, b))
Posterior <- tibble(Pr_Post = "Posterior",
                P = seq(0, 1, length.out = 200),
                a = 2,
                b = 8,
                Density = dbeta(P, a, b))
D <- bind_rows(Prior, Posterior) |> 
  mutate(Pr_Post = factor(Pr_Post, levels = c("Prior", "Posterior")))

ggplot(D, aes(P, Density, color = Pr_Post)) +
  geom_line() +
  scale_color_manual(values = c("gray50", "red"), name = NULL) +
  labs(title = "a = 2, b = 8")
```


## Carp 9: Pass

```{r}
Prior <- tibble(Pr_Post = "Prior",
                P = seq(0, 1, length.out = 200),
                a = 2,
                b = 8,
                Density = dbeta(P, a, b))
Posterior <- tibble(Pr_Post = "Posterior",
                P = seq(0, 1, length.out = 200),
                a = 3,
                b = 8,
                Density = dbeta(P, a, b))
D <- bind_rows(Prior, Posterior) |> 
  mutate(Pr_Post = factor(Pr_Post, levels = c("Prior", "Posterior")))

ggplot(D, aes(P, Density, color = Pr_Post)) +
  geom_line() +
  scale_color_manual(values = c("gray50", "red"), name = NULL) +
  labs(title = "a = 3, b = 8")
```


## All the carp

```{r}
#| echo: true
#| output-location: slide
ab <- tibble(a = rep(1, 72),
             b = rep(1, 72),
             Pass = NA)

for (ii in 2:nrow(ab)) {
  ab$Pass[ii] <- Carp$Pass[ii - 1]
  if (Carp$Pass[ii - 1] == 0) {
    ab$a[ii] <- ab$a[ii - 1]
    ab$b[ii] <- ab$b[ii - 1] + 1
  } else {
    ab$a[ii] <- ab$a[ii - 1] + 1
    ab$b[ii] <- ab$b[ii - 1]
  }
}

print(ab, n = 20)
```


## Plotting the first 16 carp

```{r}
P <- purrr::map(
  .x = 2:nrow(ab),
  .f = function(ii, ab) {
    Prior <- tibble(Pr_Post = "Prior",
                    P = seq(0, 1, length.out = 200),
                    a = ab$a[ii - 1],
                    b = ab$b[ii - 1],
                    Density = dbeta(P, a, b))
    Posterior <- tibble(Pr_Post = "Posterior",
                        P = seq(0, 1, length.out = 200),
                        a = ab$a[ii],
                        b = ab$b[ii],
                        Density = dbeta(P, a, b))
    D <- bind_rows(Prior, Posterior) |> 
      mutate(Pr_Post = factor(Pr_Post, levels = c("Prior", "Posterior")))
    
    ggplot(D, aes(P, Density, color = Pr_Post)) +
      geom_line() +
      scale_color_manual(values = c("gray50", "red"),
                         name = NULL, guide = "none") +
      labs(title = paste0("a = ", ab$a[ii], "; b = ", ab$b[ii])) +
      theme_cowplot(font_size = 10)
  },
  ab = ab)

wrap_plots(P[1:16])
```


## Final distribution of probability

```{r}
P_original <- P[[71]]
P[[71]]
```


## Shuffle the order of observations

```{r}
#| echo: true
set.seed(42364)
Carp <- Carp[sample(seq_len(nrow(Carp))), ]
Carp
```


## Repeat and plot

```{r}
ab <- tibble(a = rep(1, 72),
             b = rep(1, 72),
             Pass = NA)

for (ii in 2:nrow(ab)) {
  ab$Pass[ii] <- Carp$Pass[ii - 1]
  if (Carp$Pass[ii - 1] == 0) {
    ab$a[ii] <- ab$a[ii - 1]
    ab$b[ii] <- ab$b[ii - 1] + 1
  } else {
    ab$a[ii] <- ab$a[ii - 1] + 1
    ab$b[ii] <- ab$b[ii - 1]
  }
}

P <- purrr::map(
  .x = 2:nrow(ab),
  .f = function(ii, ab) {
    Prior <- tibble(Pr_Post = "Prior",
                    P = seq(0, 1, length.out = 200),
                    a = ab$a[ii - 1],
                    b = ab$b[ii - 1],
                    Density = dbeta(P, a, b))
    Posterior <- tibble(Pr_Post = "Posterior",
                        P = seq(0, 1, length.out = 200),
                        a = ab$a[ii],
                        b = ab$b[ii],
                        Density = dbeta(P, a, b))
    D <- bind_rows(Prior, Posterior) |> 
      mutate(Pr_Post = factor(Pr_Post, levels = c("Prior", "Posterior")))
    
    ggplot(D, aes(P, Density, color = Pr_Post)) +
      geom_line() +
      scale_color_manual(values = c("gray50", "red"),
                         name = NULL, guide = "none") +
      labs(title = paste0("a = ", ab$a[ii], "; b = ", ab$b[ii])) +
      theme_cowplot(font_size = 10)
  },
  ab = ab)

wrap_plots(P[1:16])
```


## Comparing two posteriors

```{r}
plot_grid(P_original, P[[71]])
```

The order of the data does not matter


## Informative priors

::: {.incremental}
- Almost no knowledge
  - Carp either pass or do not pass
  - a = 1; b = 1
- A little knowledge
  - We guess that the proportion is probably less than 50%
  - a = 5; b = 15
- A lot of knowledge
  - Previous studies show that <25% of carp pass
  - a = 5; b = 40
:::


## Informative priors

```{r}
P_little <- ggplot(tibble(P = seq(0, 1, length.out = 200),
                          Density = dbeta(P, 5, 15)),
                   aes(P, Density)) +
  geom_line() +
  labs(title = paste0("a = 5; b = 15")) +
  theme_cowplot(font_size = 10)

P_a_lot <- ggplot(tibble(P = seq(0, 1, length.out = 200),
                         Density = dbeta(P, 5, 40)),
                  aes(P, Density)) +
  geom_line() +
  labs(title = paste0("a = 5; b = 40")) +
  theme_cowplot(font_size = 10)

plot_grid(P_little, P_a_lot)
```


## Posteriors from informative priors

Different priors = Different posteriors

```{r}
P_little <- ggplot(tibble(P = seq(0, 1, length.out = 200),
                          Density = dbeta(P, 5 + 9, 15 + 62)),
                   aes(P, Density)) +
  geom_line() +
  labs(title = paste0("a = 14; b = 77")) +
  theme_cowplot(font_size = 10)

P_a_lot <- ggplot(tibble(P = seq(0, 1, length.out = 200),
                         Density = dbeta(P, 5 + 9, 40 + 62)),
                  aes(P, Density)) +
  geom_line() +
  labs(title = paste0("a = 14; b = 102")) +
  theme_cowplot(font_size = 10)

plot_grid(P_little, P_a_lot)
```


## Frequentist confidence intervals [see @Morey2016-yh]

- 95% CI for a proportion of 9 / 72 is 0.07 - 0.22 (via Wilson's method)
- "If we repeated this experiment over and over, 95% of those experiments would have a proportion between 0.07	and 0.22."

*Not*:

- 95% certain that the proportion is between 0.07 and 0.22
- 95% probability that the proportion is between 0.07 and 0.22


## Bayesian highest density interval

- 95% highest density interval (HDI) for the proportion (given a Beta(1, 1) prior and the data) is 0.06 to 0.22

"There is a 95% probability that the true proportion falls between 0.06 and 0.22".


## Advantages of Bayesian inference 1

- Randomness or noise is a property of data not of sampling
  - Is there one single population body mass of rabbits in a field?
- Include prior knowledge in the model
- Posteriors represent the relative plausibility of estimates (given a model, data, and priors)
- Posteriors can be interpreted as probabilities


## Advantages of Bayesian inference 2

- Regularizing priors can make complex models feasible
  - Multilevel/mixed models
  - Generalized linear models
  - Nonlinear models


## References

::: {#refs}
:::

