---
title: "Untitled"
format: html
---

## Maximum likelihood

```{r}
set.seed(34279)
D <- tibble(x = rnorm(40, 47, 5))

Probs <- tibble(
  Pred = c("Mean = 45; sd = 5",
           "Mean = 47; sd = 3",
           "Mean = 47; sd = 6",
           "Mean = 42; sd = 10",
           "Mean = 40, sd = 3"),
  Mean = c(45, 47, 47, 42, 40),
  SD = c(5, 3, 6, 10, 2)
) |> 
  mutate(Pred = fct_inorder(Pred))


Dens <- map_dfr(
  .x = 1:nrow(Probs),
  .f = function(ii, Probs){
    tibble(Label = Probs$Pred[ii],
           x = seq(30, 65, length.out = 200),
           y = dnorm(x, mean = Probs$Mean[ii], sd = Probs$SD[ii]))
  },
  Probs = Probs)

ggplot(D, aes(x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30) +
  geom_line(data = Dens,
            aes(x, y, color = Label),
            linewidth = 2) +
  scale_color_brewer(type = "qual", name = NULL) +
  labs(x = "Measure", y = "Density") +
  theme(legend.position = c(0.7, 0.8))
```


## The analytical answers

```{r}
#| echo: true
mean(D$x)
sd(D$x)
```


## Normal (Gaussian) probability density:

$$f(x) = \frac{1}{\sigma \sqrt{2\pi} } e^{-\frac{1}{2}\left(\frac{x-\mu}{\sigma}\right)^2}$$

where

- $\mu$ is the mean
- $\sigma$ is the standard deviation
- $\pi$ and $e$ are constants


## A function for the normal probability density

```{r}
#| echo: true
d_norm <- function(Theta, x, log_scale = TRUE) {
  mu <- Theta[1]
  sigma <- Theta[2]

  p <- 1 / (sigma * sqrt(2 * pi)) * 
    exp(-1 / 2 * ((x - mu) / sigma) ^ 2)
  
  if (log_scale) p <- log(p)
  
  return(p)
}
```


## Testing the function

```{r}
#| echo: true
d_norm(Theta = c(0, 1), x = 1)
d_norm(Theta = c(0, 1), x = -1:2)
d_norm(Theta = c(0, 1), x = 10, log_scale = FALSE)
d_norm(Theta = c(0, 1), x = 10, log_scale = TRUE)
```


## Maximum likelihood estimation

Function to return the model likelihood:

```{r}
#| echo: true

Log_Lik <- function(Theta, x) {
  sum(d_norm(Theta, x))
}
```

Optimize `Log_Lik()`:

```{r}
#| echo: true
#| output-location: slide
optim(c(40, 1), # Start at 40, 1
      Log_Lik,
      x = D$x,
      control = list(fnscale = -1,      # Maximize
                     reltol = 10^-100)) # Stopping point

mean(D$x)
sd(D$x)
```


## What's the problem?

Sample standard deviation (unbiased)

$$s = \sqrt{\frac{1}{N-1}\sum_{i}(x-\bar x)^2}$$

MLE estimate of standard deviation (biased)

$$\hat\sigma=\sqrt{\frac{1}{N}\sum_{i=1}^N(x-\bar x)^2}$$


## With `glm()`

- `~ 1` tells `glm()` to fit only an intercept
- `"gaussian"` is the default, so we don't need to include it

```{r}
#| echo: true
fm <- glm(D$x ~ 1, family = "gaussian")
```


## `glm()` summary and log-likelihood

```{r}
#| echo: true
summary(fm)
logLik(fm)
```
