---
title: "Distributions and Null Models"
subtitle: ""
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
---

## Monte Carlo Methods

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(cowplot)
ggplot2::theme_set(theme_cowplot())
```

"Monte Carlo Methods" - a general broad term that includes all methods that involve any random sampling. 

Named for the famous casino - how are we doing images?

## Random Sampling

- Values from a distribution (e.g. a normal distribution)
- Values from a dataset
- Values from a set of possible parameters
- A random order for a set of values

```{r, echo=TRUE}

rand.vals <- tibble("Random" = rnorm(1000))

ggplot(rand.vals, aes(Random)) +
  geom_histogram(bins=100)  

```

## Why would you need Monte Carlo Methods?

## Your Dataset Violates the Assumption(s) of Parametric Tests

## You want to ask how generalizable your results are

## Your question isn't answered by a typical statistical test

- Groups can be different due to multiple causes only some of which are of interest

## There is not a standard way to estimate a confidence interval









