---
title: "Foundations of Sampling"
subtitle: "Randomization"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
code-annotations: hover
---

## Sampling from data sets: foundations

- Jackknife
- Bootstrap
- Randomization

## What is randomization?

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(cowplot)
library(readxl)
library(viridis)
ggplot2::theme_set(theme_cowplot())
set.seed(487622)
```

- A technique for generating an empirical null distribution by repeatedly putting the data in a random order
- Primarily for hypothesis testing
    - Wide applicability
    - Limited to the focal hypothesis being tested

**Assumption:** Under the null hypothesis, observations are random draws from a common population


## The General Procedure

1. Decide on a test statistic
1. Calculate the test statistic for the *observed* data
1. Randomly shuffle the observations
1. Calculate the test statistic for that group
1. Repeat many times to generate an empirical null distribution
1. Determine the proportion of random combinations resulting in a test statistic more extreme than the observed value ("empirical *P*")

## Are male and female jackal mandible lengths different?

```{r}
#| echo: false

M <- read_excel("Data/Jackals.xlsx")
M <- M |> mutate(Sex = factor(Sex))

M.set |> 
  ggplot(aes(Mandible, fill = Sex)) + 
  geom_dotplot(method = "histodot", binwidth = 0.5, stackgroups = TRUE) +
  scale_fill_viridis_d() +
  scale_y_continuous(NULL, breaks = NULL) 

M.set |>
  ggplot(aes(Mandible)) + 
  geom_dotplot(method = "histodot", binwidth = 0.5, stackgroups = TRUE) +
  scale_color_viridis_d() +
  scale_y_continuous(NULL, breaks = NULL) 

M.set$Sex <- sample(M.set$Sex)

M.set |>
  ggplot(aes(Mandible, fill = Sex)) + 
  geom_dotplot(method = "histodot", binwidth = 0.5, stackgroups = TRUE) +
  scale_fill_viridis_d() +
  scale_y_continuous(NULL, breaks = NULL) 
```


## General Considerations

- Bootstrap methods assume your observed dataset approximates the true distribution
    - This assumption is not likely for small samples
- Every use case must be shown to be valid via simulation    
- Decisions
    - How many samples?
    - What CI method?
    - How to simulate data to test the method?


## Common Use Cases

- Confidence intervals (most common)
- Hypothesis tests (less common)
    - Support for branches on phylogenetic trees (Felsenstein, J. 1985. Confidence Limits on Phylogenies: An Approach Using the Bootstrap. *Evolution* 39:783–791.)

