---
title: "Genetic algorithm"
author:
  - Elizabeth King
  - Kevin Middleton
format:
  revealjs:
    theme: [default, custom.scss]
    standalone: true
    self-contained: true
    logo: QMLS_Logo.png
    slide-number: true
    show-slide-number: all
code-annotations: hover
bibliography: Randomization.bib
csl: evolution.csl
---


## Sampling for parameter estimation

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(cowplot)
library(rgenoud)

ggplot2::theme_set(theme_cowplot(font_size = 18))
set.seed(6452730)
```

Many problems are more easily (or possibly) solved by not using a model with a traditional likelihood

1. Sample from distributions and compare to observed values (Approximate Bayesian Computation)
2. Generate "populations" of parameters and let them "evolve" across generations (Genetic Algorithm)


## Genetic algorithm

- "Genetic" in name only
    - Inspired by ideas from evolution (populations, selection, mutation, crossover)
- General optimization approach
    - "Stochastic search algorithm" [@Scrucca2013-jf]
    - No likelihood needed


## Genetic algorithm history

- Introduced / synthesized: [@Holland1975-nz; @Goldberg1989-zk]
- Modern implementation with derivatives: [@Sekhon1998-hs; @Mebane2011-vy]


## Benefits


## R packages for GA

1. [`rgenoud`](http://sekhon.berkeley.edu/rgenoud)
2. `GA` [@Scrucca2013-jf]


## Simulate data

```{r}
#| echo: false

set.seed(234798)
y <- rgamma(500, shape = 3, rate = 3)

ggplot(tibble(y), aes(y)) +
  geom_histogram(bins = 30)
```


## Model statement and function

$$y \sim Gamma(shape, rate)$$

```{r}
#| echo: true

gamma_sim <- function(p) {
  log_lik <- sum(dgamma(y, p[1], p[2], log = TRUE))
  return(log_lik)
}

gamma_sim(c(3, 3))
```


## Choosing the parameter "search" space

- What are the likely values for the parameters?
    - What distributions?
    - What limits?
- Similar to choosing Bayesian priors
- Uniform priors will work fine


## `genoud()` options: `P1` - `P9`


## Fitting the model

```{r}
genoud(gamma_sim,
       nvars = 2,
       max = TRUE,
       Domains = matrix(c(0, 10, 
                          0, 10), nrow = 2, byrow = TRUE),
       pop.size = 5000)
```


## Differences from ABC

- No prior distributions to draw from
- No samples to process


## Challenging models

- "Intractable" likelihoods
    - Flat or oddly shaped surfaces
- Nonlinear models
- Multiple optima


## Non-linear model: Gompertz growth equation

$$y(t) = a e^{-b e^{-c t}}$$

- *a* is the asymptotic size
- *b* is the $x$ displacement of the entire curve
- *c* is growth rate

```{r}
#| echo: true

Gompertz <- function(t, a, b, c) {
  a * exp(-b * exp(-c * t))
}

```


## Testing

```{r}
#| echo: false

GG <- tibble(t = seq(0, 25, length.out = 200),
             y = Gompertz(t, a = 5, b = 1, c = 0.5))

ggplot(GG, aes(t, y)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(limits = c(0, 6))
```


## Simulate data

```{r}
#| echo: false

set.seed(436578)
GGsim <- tibble(t = runif(20, min = 0, max = 25),
                y = Gompertz(t, a = 5, b = 1, c = 0.5) +
                  rnorm(20, 0, 0.25))

ggplot() +
  geom_line(data = GG, aes(t, y), linewidth = 1) +
  geom_point(data = GGsim, aes(t, y), size = 3) +
  scale_y_continuous(limits = c(0, 6))
```


## Setting up `genoud()`

```{r}
Gomp_MSE <- function(p, GGsim) {
  pred <- Gompertz(GGsim$t, p[1], p[2], p[3])
  obs <- GGsim$y
  return(mean((obs - pred) ^ 2))
}

gen <- genoud(fn = Gomp_MSE,
              GGsim = GGsim,
              nvars = 3,
              pop.size = 20000)
```


## Visualizing the model

```{r}
ggplot() +
  geom_line(data = GG, aes(t, y), linewidth = 1) +
  geom_line(
    data = tibble(t = seq(0, 25, length.out = 200),
                  y = Gompertz(t, a = gen$par[1], b = gen$par[2],
                               c = gen$par[3])),
    aes(t, y),
    color = "red", linewidth = 1) +
  geom_point(data = GGsim, aes(t, y), size = 3) +
  scale_y_continuous(limits = c(0, 6))
```


## References

::: {#refs}
:::

