---
title: "Problem Set 4"
author:
  - Your Name Here
format: 
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    toc-title: Contents
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(cowplot)
library(brms)
library(priorsense)
library(tidybayes)
library(distributional)
library(modelr)

options(brms.backend = "cmdstan")
theme_set(theme_cowplot(font_size = 10))

# Required files for this problem set:
# remotes::install_github("n-kall/priorsense")
```


## `priorsense` Package

Kallioinen, N., T. Paananen, P.-C. Bürkner, and A. Vehtari. 2021. Detecting and diagnosing prior and likelihood sensitivity with power-scaling.

`remotes::install_github("n-kall/priorsense")`


```{r}
EW <- read_csv("../Data/Earwigs.csv", show_col_types = FALSE)

ggplot() +
  geom_point(data = EW, aes(Density, Proportion_forceps),
             color = "steelblue", size = 3)
```

Models

```{r}
get_prior(Proportion_forceps ~ 1 + Density,
           data = EW)

fm <- brm(Proportion_forceps ~ 1 + Density,
          data = EW,
          prior = c(prior(normal(0, 0.1), class = b),
                    prior(normal(0, 1), class = Intercept),
                    prior(normal(0, 1), class = sigma)),
          refresh = 0)

prior_summary(fm)

powerscale_sensitivity(fm)

powerscale_plot_quantities(
  powerscale_sequence(fm),
  quantities = c("mean", "sd"),
  variables = c("b_Density", "b_Intercept")
)
```

Working with tidybayes: http://mjskay.github.io/tidybayes/articles/tidy-brms.html

```{r}
library(tidybayes)
fm |> tidy_draws() |> summarise_draws()

fm |>
  spread_rvars(b_Intercept, b_Density, sigma) |> 
  median_hdi(.width = 0.89)

fm |>
  spread_rvars(b_Intercept, b_Density, sigma) |> 
  pivot_longer(cols = everything()) |> 
  ggplot(aes(y = name, dist = value)) +
  stat_pointinterval(.width = c(0.89, 0.97))

library(modelr)
EW |>
  data_grid(Density = seq_range(Density, n = 100)) |>
  add_epred_draws(fm) |> 
  ggplot(aes(x = Density, y = Proportion_forceps)) +
  stat_lineribbon(aes(y = .epred), alpha = 0.5) +
  geom_point(data = EW)
```


## ANOVA-like

- Load data
- `Survival` is measured in days
- `Mismatch` to factor: low, medium, high (not not default alphabetical ordering)

```{r}
HT <- read_csv("../Data/Heart_Transplants.csv", show_col_types = FALSE) |> 
  mutate(Mismatch = fct_inorder(Mismatch))

ggplot(HT, aes(x = Mismatch, y = Survival)) +
  geom_point(position = position_jitter(width = 0.1, seed = 4564356)) +
  stat_summary(fun = mean, geom = "point", size = 3, color = "red") +
  stat_summary(fun.data = mean_se, geom = "errorbar", 
               width = 0.1,
               linewidth = 1,
               color = "red")

HT |> 
  group_by(Mismatch) |> 
  summarize(mean_Survival = mean(Survival),
            sd_Survival = sd(Survival))

HT <- HT |> 
  mutate(logSurvival = log(Survival))

ggplot(HT, aes(x = Mismatch, y = logSurvival)) +
  geom_point(position = position_jitter(width = 0.1, seed = 4564356)) +
  stat_summary(fun = mean, geom = "point", size = 3, color = "red") +
  stat_summary(fun.data = mean_se, geom = "errorbar", 
               width = 0.1,
               linewidth = 1,
               color = "red")
```


```{r}
get_prior(logSurvival ~ Mismatch - 1,
          data = HT)

PP <- brm(logSurvival ~ Mismatch - 1,
          data = HT,
          prior = c(prior(normal(4, 3), class = b)),
          refresh = 0,
          sample_prior = "only")

pp_check(PP, type = "stat_grouped", group = "Mismatch",
         stat = "mean",
         ndraws = 500,
         binwidth = 0.5)
```

```{r}
fm <- brm(logSurvival ~ Mismatch - 1,
          data = HT,
          prior = c(prior(normal(4, 3), class = b)),
          refresh = 0)

prior_summary(fm)

powerscale_sensitivity(fm)

powerscale_plot_quantities(
  powerscale_sequence(fm),
  quantities = c("mean", "sd"),
  variables = c("b_MismatchLow", "b_MismatchMedium", "b_MismatchHigh")
)

summary(fm)
```

```{r}
pp_check(fm, type = "stat_grouped", group = "Mismatch",
         stat = "mean",
         ndraws = 500)
```

```{r}
fm |>
  spread_rvars(b_MismatchLow, b_MismatchMedium, b_MismatchHigh) |> 
  set_names(distinct(HT, Mismatch) |> pull()) |> 
  median_hdi(.width = 0.89)

fm |>
  spread_rvars(b_MismatchLow, b_MismatchMedium, b_MismatchHigh) |> 
  set_names(distinct(HT, Mismatch) |> pull()) |> 
  pivot_longer(cols = everything()) |> 
  mutate(name = fct_inorder(name)) |> 
  ggplot(aes(y = name, dist = value)) +
  stat_pointinterval(.width = c(0.89, 0.97))

HT |> 
  data_grid(Mismatch) |> 
  add_predicted_draws(fm) |> 
  ggplot(aes(x = .prediction, y = Mismatch)) +
  stat_slab()

HT |> 
  data_grid(Mismatch) |> 
  add_predicted_draws(fm) |> 
  ggplot(aes(x = .prediction, y = Mismatch)) +
  stat_interval(.width = c(0.50, 0.89, 0.97)) +
  geom_point(aes(x = logSurvival), data = HT) +
  scale_color_brewer()


library(distributional)
HT |> 
  data_grid(Mismatch) |> 
  add_epred_draws(fm, dpar = c("mu", "sigma")) |> 
  sample_draws(30) |> 
  ggplot(aes(y = Mismatch)) +
  stat_slab(aes(xdist = dist_normal(mu = mu, sigma = sigma)), 
            slab_color = "gray65", alpha = 0.1, fill = NA) +
  geom_point(aes(x = logSurvival), data = HT, shape = 21, fill = "#9ECAE1", size = 3)
```

```{r}
fm |>
  spread_rvars(b_MismatchLow, b_MismatchMedium, b_MismatchHigh) |> 
  set_names(distinct(HT, Mismatch) |> pull()) |> 
  mutate(Med_v_Low = Medium - Low,
         High_v_Low = High - Low,
         .keep = "none") |> 
  pivot_longer(cols = everything()) |> 
  ggplot(aes(y = name, dist = value)) +
  stat_pointinterval(.width = c(0.5, 0.89))
```


## 2x2 factorial design

The file `Bird_Plasma.xlsx` contains factorial data on blood plasma calcium concentration (`Calcium`, mg/100 mL) in male and female birds (`Sex`) each of which was treated or not with a hormone (`Treatment`).

Load the data, and convert hormone and sex to factors. The levels of `Treatment` are "Hormone" and "None". Relevel `Treatmeant` so that "None" the base level.

```{r, message = FALSE}
# FIXME
BP <- readxl::read_excel("../Data/Bird_Plasma.xlsx") |> 
  mutate(Treatment = factor(Treatment),
         Sex = factor(Sex),
         Treatment = fct_relevel(Treatment, "None"))

BP |> count(Treatment, Sex)

ggplot(BP, aes(x = Sex,
               y = Calcium,
               color = Treatment,
               group = Treatment)) +
  geom_point(position = position_jitter(width = 0.05, seed = 474577),
             size = 3) +
  stat_summary(fun = mean, geom = "point", pch = 5, size = 5) +
  stat_summary(fun = mean, geom = "line") +
  scale_color_manual(values = c("brown", "darkgreen"))
```

Only 5 points per group, so the prior is potentially very powerful.

```{r}

BP <- BP |> 
  mutate(Sex_Trt = paste(Sex, Treatment, sep = "_"))

get_prior(Calcium ~ Sex_Trt - 1,
          data = BP)

PP <- brm(Calcium ~ Sex_Trt - 1,
          data = BP,
          prior = c(prior(normal(20, 15), class = b),
                    prior(normal(0, 10), class = sigma)),
          refresh = 0,
          sample_prior = "only")

pp_check(PP, type = "stat_grouped", group = "Sex_Trt",
         stat = "mean",
         ndraws = 500)

```


```{r}
fm <- brm(Calcium ~ Sex_Trt - 1,
          data = BP,
          prior = c(prior(normal(20, 15), class = b),
                    prior(normal(0, 10), class = sigma)),
          refresh = 0)

prior_summary(fm)

powerscale_sensitivity(fm)

powerscale_plot_quantities(
  powerscale_sequence(fm),
  quantities = c("mean", "sd"),
  variables = c("b_Sex_TrtFemale_Hormone", "b_Sex_TrtFemale_None"))
powerscale_plot_quantities(
  powerscale_sequence(fm),
  quantities = c("mean", "sd"),
  variables = c("b_Sex_TrtMale_Hormone", "b_Sex_TrtMale_None"))
powerscale_plot_quantities(
  powerscale_sequence(fm),
  quantities = c("mean", "sd"),
  variables = c("sigma"))

summary(fm)

BP |> 
  data_grid(Sex_Trt) |> 
  add_predicted_draws(fm) |> 
  ggplot(aes(x = .prediction, y = Sex_Trt)) +
  stat_slab(alpha = 0.5, fill = "firebrick4")


BP |> 
  data_grid(Sex_Trt) |> 
  add_predicted_draws(fm) |> 
  median_hdi(width = 0.89)
```

Contrasts

```{r}
post <- as_draws_df(fm) |> 
  as_tibble() |> 
  select(starts_with("b_")) |> 
  mutate(`F: Horm. vs. C.` = b_Sex_TrtMale_Hormone - b_Sex_TrtMale_None,
         `M: Horm. vs. C.` = b_Sex_TrtFemale_Hormone - b_Sex_TrtFemale_None,
         .keep = "none")

post |> 
  pivot_longer(cols = everything()) |> 
  ggplot(aes(value, color = name, fill = name)) +
  geom_density(alpha = 0.5)

median_hdi(post, .width = 0.89)
```


## Multiple continuous predictors

Tradeoff between fat and lactose in mammal milk.

Example for multiple regression

```{r}
#| warning: false
MM <- readxl::read_excel("../Data/Milk.xlsx") |> 
  select(kcal.per.g, perc.fat, perc.lactose) |> 
  drop_na() |> 
  rename(Milk_energy = kcal.per.g,
         Fat = perc.fat,
         Lactose = perc.lactose)

p1 <- plot_grid(ggplot(MM, aes(Fat, Milk_energy)) + geom_point(),
                ggplot(MM, aes(Lactose, Milk_energy)) + geom_point(),
                ncol = 2)
p1
```

```{r}
PP <- brm(Milk_energy ~ Fat + Lactose,
          data = MM,
          prior = prior(normal(0, 0.05), class = b),
          sample_prior = "only",
          refresh = 0)

pp_check(PP, ndraws = 100)

pp_check(PP, type = "stat", stat = "median")
```

```{r}
fm <- brm(Milk_energy ~ Fat + Lactose,
          data = MM,
          prior = prior(normal(0, 0.05), class = b),
          refresh = 0)

prior_summary(fm)

powerscale_sensitivity(fm)

powerscale_plot_quantities(
  powerscale_sequence(fm),
  quantities = c("mean", "sd"),
  variables = c("b_Intercept", "b_Fat", "b_Lactose"))

pred_values <- crossing(
  Fat = seq(3.9, 55.5, length.out = 50),
  Lactose = c(30, 50, 70)
)

p_pred <- posterior_epred(fm, newdata = pred_values)

pred_values <- pred_values |> 
  mutate(Q50 = apply(p_pred, MARGIN = 2, FUN = quantile, prob = 0.5),
         Q5.5 = apply(p_pred, MARGIN = 2, FUN = quantile, prob = 0.055),
         Q94.5 = apply(p_pred, MARGIN = 2, FUN = quantile, prob = 0.945),
         Lactose = factor(Lactose))
pred_values

p2 <- ggplot() +
  geom_ribbon(data = pred_values,
              aes(x = Fat, ymin = Q5.5, ymax = Q94.5,
                  fill = Lactose), alpha = 0.25) +
  geom_line(data = pred_values,
            aes(x = Fat, y = Q50, color = Lactose)) +
  facet_grid(. ~ Lactose)
plot_grid(p1, p2, nrow = 2)
```

